---
title: "False Positives"
author: "Pietro Franceschi"
date: "4 February 2019"
output: rmarkdown::github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  fig.path = "figs/falsepos"
)
```

This first demo deals with "random organization". Asw we discussed in the lecture, an experiment is always performed on a subset of the population, so a measured "organization" 

* can mirror a population level characteristics
* can be the result of a "unlucky" sampling

To undertand the second point, just consider an hypothetic sampling of the human population made by an extraterrestrial entity. If he random sample a couple of humans, it will likely conclude that we are all chinese! 

Let's make a more scientific example. Let's create a fake proteomics experiment where I'm measuring the concentration of 1000 proteins on a group of 20 samples. And let's suppose that I want to study the correlation between the different proteins to infer new regulatory networks ...

To simulate a dummy experiment I will fill my expression matrix with random numbers ...

```{r cars}
## Random expression data matrix

DM <- matrix(rnorm(20000), nrow = 20)
dim(DM)

```

That's the data matrix. Is there true organization there? NO!

Let's calculate the correlation among the different proteins ...

```{r}
mycorr <- cor(DM)
dim(mycorr)
```

Now we get only the upper triangle of the matrix and we look to the distribution of the correlation coefficients

```{r}
hist(mycorr[upper.tri(mycorr)], 
     col = "steelblue", 
     breaks = 100,
     main = "Random Correlations")


```

Unexpectedly ( ;-) ) ... we also have high correlation variables!
The code below gives you the indices of the variables showing high correlation, excluding the 1s


```{r}
highcorrid <- which(mycorr > 0.8 & mycorr != 1, arr.ind = TRUE)
highcorrid
```


```{r}
plot(DM[,highcorrid[1,1]],DM[,highcorrid[1,2]], col = "darkred", pch = 19, cex = 2)
```


Here the presence of high correlation is clear!!!
But this correlation does not have a "biological" origin ... so it does not represent a truly scientific result ...
